---
title: "Gov51-PS4: Problem Set 4: Instrumental Variables"
author: "Mizuki Yashiro"
date: "2026-04-10"
output: pdf_document
---
Github: https://github.com/mizukiyashiro/gov51-ps4

# 1. Setup and Exploration

## 1.1 Load packages and data
```{r}
library(tidyverse)
library(estimatr)
library(modelsummary)
```

```{r}
card <- read_csv("data/card1995.csv")
```

## 1.2 Describe the data

```{r}
nrow(card)
```

The dataset contains 3010 observations. The unit of observation is an individual.

## 1.3 Distribution of years of education

```{r}
ggplot(card, aes(x = educ)) +
  geom_histogram(binwidth = 1)
```
```{r}
mean(card$educ, na.rm = TRUE)
```

The mean years of education is 13.26346. The histogram shows spikes around 12-13 years and at 16 years. These spikes likely reflect common educational milestones. 12 years corresponds to high school graduation, while 16 years of education corresponds to completing a four-year college degree. The concentration of observations at these values leads me to believe that individuals tend to cluster at standard credentialed levels of graduation rather than having a smooth distribution of years of schooling.

## 1.4 The naive relationship

```{r}
ggplot(card, aes(x = educ, y = lwage)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE)
```

The scatter plot shows a positive relationship between education and log weekly wages. As years of education increase, log wages tend to be higher on average. This is reflected in the upward-sloping regression line, indicating that individuals with more education generally earn higher wages. However, there is  variation in wages at each level of education, suggesting that education is not the only factor influencing earnings.

## 1.5. Why OLS might be misleading

Having years of education does not tell you about the characteristics of the people themselves. People who already have high innate abilities inherently might complete more education. One reason is ability bias. People with higher innate ability may both get more education and earn higher wages, even if education itself does not raise wages. A second reason is family background. People from wealthier or more supportive families may be able to stay in school longer and also have access to better job opportunities later on. Because these factors affect both education and wages, the OLS estimate may overstate the causal effect of education.

# 2. The Instrument: College Proximity

## 2.1 What the instrument measures

```{r}
mean(card$nearc4, na.rm = TRUE)
```

The share of respondents who grew up near a 4-year college is 68.20598%.

## 2.2 Relevance: does proximity affect education?

```{r}
first_stage <- lm_robust(
  educ ~ nearc4 + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2")

summary(first_stage)
```

The coefficient on nearc4 is 0.32711 with a SE = 0.080554 and t-statistic = 4.0607. This means that growing up near a 4-year college is associated with 0.32711 additional years of education, holding other factors constant. The positive and statistically significant coefficient suggests that proximity to a college increases educational attainment, likely because it reduces the financial and logistical costs of attending college.

## 2.3 Checking instrument strength

```{r}
t_stat <- coef(summary(first_stage))["nearc4", "t value"]
f_stat <- t_stat^2

t_stat
f_stat
```

The t-statistic on nearc4 is 4.0607, so the first-stage F-statistic is 16.48928449. Since this exceeds 10, the instrument is considered strong according to the rule of thumb. Instrument strength matters because weak instruments can produce biased and imprecise 2SLS estimates.

## 2.4

(A) The exclusion restriction requires that, after controlling for education and the other covariates, growing up near a 4-year college affects log wages only through its effect on education and not through any other channel. Being clsoe to colelge is the IV.

(B) One possible violation is that people who grew up near a college may also have grown up in places with stronger local labor markets, better public schools, or more urban economic opportunities. In that case, nearc4 could affect wages directly, not only through schooling, which would violate the exclusion restriction. beiong in proximity to college === more educational opportuntiies

## 2.5 Balance check

```{r}
card |>
  group_by(nearc4) |>
  summarize(
    mean_black = mean(black, na.rm = TRUE),
    mean_smsa = mean(smsa, na.rm = TRUE),
    mean_south = mean(south, na.rm = TRUE))
```

The groups are not balanced on pre-treatment characteristics. Individuals who grew up near a 4-year college are less likely to be Black (0.2118850 vs. 0.2800418), much more likely to live in a metropolitan area (0.8222114 vs. 0.4785789), and less likely to live in the South (0.3292742 vs. 0.5632184). These differences lead me to believe that proximity to a college is correlated with background characteristics, meaning that the instrument is not randomly assigned. Individuals near colleges are more likely to be in urban, non-Southern areas, which may have different economic conditions. This is concerning because it means that nearc4 could be related to wages through variables other than education. However, including these variables as controls helps mitigate this concern.

# 3. The Wald Estimator


## 3.1 Compute the reduced form

```{r}
reduced_form <- lm_robust(
  lwage ~ nearc4 + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2")

summary(reduced_form)
```

The coefficient on nearc4 is 0.040854, with a SE = 0.0160993, and t = 2.538. This means that growing up near a 4-year college is associated with an increase of 0.040854 higher log weekly wages than those who did not, holding other factors constant. Because the dependent variable is in logs, this corresponds to approximately a 100 x 0.040854% increase in wages. This estimate is the reduced-form effect of the instrument on wages, combining both the effect of proximity on education and the effect of education on wages. In other words, it reflects the overall relationship between growing up near a college and earnings, without yet isolating the causal effect of education itself.

## 3.2 Calculate the Wald estimator by hand

```{r}
rf <- 0.040854
fs <- 0.32711
wald <- rf / fs

wald
```

The Wald estimator by hand is 0.1248938.

## 3.3 Interpret the Wald estimator

The Wald estimate implies that for individuals whose education was affected by growing up near a 4-year college, one additional year of education increases log weekly wages by 0.1248938. This corresponds to approximately a 100 x 0.1248938% increase in wages.

# 4. OLS vs. 2SLS

## 4.1 OLS estimate

```{r}
ols <- lm_robust(
  lwage ~ educ + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2")

summary(ols)
```

The OLS coefficient on education is 0.071498.

## 4.2 2SLS estimate

```{r}
iv <- iv_robust(
  lwage ~ educ + black + smsa + south + exper + expersq + married |
    nearc4 + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2")

summary(iv)
```

The 2SLS coefficient on education is 0.124893. This matches the Wald estimate from Section 3.2. This is expected because in the case of a single instrument, the 2SLS estimator is numerically identical to the Wald estimator. Both approaches use variation in education induced by the instrument 'nearc4' to estimate the causal effect of education on wages.

## 4.3 Side-by-side comparison

```{r}
modelsummary(
  list("OLS" = ols, "2SLS" = iv),
  coef_map = c("educ" = "Years of Education"),
  gof_map = c("nobs", "r.squared"),
  output = "markdown")
```

## 4.4 Why do OLS and 2SLS differ?

(A) The 2SLS estimate, 0.124893 is larger than the OLS estimate, 0.071498.

(B) One reason OLS may underestimate the return to education is measurement error in education, which biases OLS estimates toward zero. Another reason is that the IV estimate captures the return for compliers. These are individuals whose education increased due to proximity to a college, and thus may have higher marginal returns to education than the average person. These factors can lead the 2SLS estimate to exceed the OLS estimate.

(C) The 2SLS estimate is more credible as a causal estimate because it uses exogenous variation in education induced by proximity to a college. This helps isolate the effect of education from confounding factors such as ability or family background, which may bias the OLS estimate.

## 4.5 Adding IQ as a control

```{r}
ols_iq <- lm_robust(
  lwage ~ educ + IQ + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2")

summary(ols_iq)
```
```{r}
iv_iq <- iv_robust(
  lwage ~ educ + IQ + black + smsa + south + exper + expersq + married |
    nearc4 + IQ + black + smsa + south + exper + expersq + married,
  data = card,
  se_type = "HC2")

summary(iv_iq)
```

(A) When IQ is added as a control, the OLS coefficient on education decreases from 0.071498 to 0.06757. This means that part of the positive relationship between education and wages in the OLS regression was driven by omitted ability bias, since IQ is positively correlated with both education and wages.

(B) When IQ is added, the 2SLS estimate decreases from 0.124893 to 0.1093899. This represents a modest change relative to the OLS case.

(C) The comparison suggests that omitted ability bias plays a larger role in OLS than in 2SLS. Including IQ reduces the OLS estimate noticeably, indicating that ability was confounding the relationship between education and wages. In contrast, the 2SLS estimate changes less when IQ is added, suggesting that the instrumental variable approach already accounts for unobserved ability to a greater extent.

# 5. Interpretation and LATE

## 5.1 Who are the compliers?

The compliers are individuals whose educational attainment increased because they grew up near a 4-year college. These are people who would not have obtained as much education if they had lived farther away, but did pursue additional schooling because proximity reduced the cost or difficulty of attending college.

## 5.2 Write the LATE interpretation

The 2SLS estimate measures the causal effect of an additional year of education on log weekly wages for individuals whose schooling increased because they grew up near a 4-year college, using geographic variation in college proximity as the source of exogenous variation.

## 5.3 External validity

The LATE is not necessarily equal to the average treatment effect (ATE) for the full population. The estimate applies specifically to compliers—individuals whose education decisions are influenced by proximity to a college. These individuals may have different returns to education than the average person. For example, compliers may have higher marginal returns to education if a small reduction in cost was enough to induce them to acquire more schooling. Therefore, the IV estimate may not generalize to everyone in the sample.

